---
title: "StateCommute: from pathway scores to per-state representative pathways"
author: "StateCommute"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{StateCommute walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
library(StateCommute)
```

## Overview

StateCommute characterizes disease-associated cell states at the **pathway**
level. The workflow has two complementary halves that meet at a bridge:

1. **Cell side (needs a Seurat object).** Score gene sets per cell with irGSEA
   (AUCell), factorize the resulting pathway × cell matrix with NMF to recover
   functional modules, and cluster cells into states.
2. **Gene side (pure gene-set math).** Collapse redundant pathways by gene-set
   overlap similarity and pick one representative per redundancy group.

The two grouping logics are different on purpose. **NMF groups pathways by how
they co-vary across cells; overlap reduction groups them by shared genes.** The
bridge, `reduce_pathways_for_states()`, runs the overlap reduction *within* each
state's NMF-selected pathways, using the NMF loading as the tie-breaker for which
pathway represents the group.

This vignette shows the cell-side stages as (non-evaluated) code you would run on
your own data, then demonstrates the gene-side reduction engine on a small
simulated gene-set collection so every step below actually runs.

## Part 1 — Cell side (on your Seurat object)

These stages require a Seurat object plus `irGSEA`, `RcppML`, `msigdbr` and
`Seurat`. They are shown for reference and are not evaluated here.

### Gene sets

```{r genesets, eval = FALSE}
gs <- prepare_genesets_msigdb(
  species    = "mouse",
  db_species = "MM"
  # default specs: MH, M2:CP:REACTOME, M2:CP:WIKIPATHWAYS, M7, M5:GO:BP (size 20-200)
)

# Object-aware: keep only genes present in the data, then a size floor.
gs <- filter_genesets_to_features(gs, rownames(obj[["RNA"]]))
gs <- standardize_genesets(gs, min_geneset_size = 5)
```

### Scoring and NMF

Only AUCell is used downstream because NMF needs a non-negative matrix.

```{r scoring-nmf, eval = FALSE}
obj <- score_pathways_irgsea(obj, gs, methods = "AUCell", ncores = 12)

mat <- get_pathway_cell_matrix(obj)     # pathways x cells (AUCell)
nm  <- run_nmf(mat, k = 20, seed = 1234)
obj <- embed_nmf(obj, nm)               # adds an "nmf" DimReduc
```

### Clustering and (optional) QC

```{r cluster-qc, eval = FALSE}
obj <- cluster_states(obj, resolution = 0.3)   # SNN on NMF + nmf_umap

# Optional ambient-contamination QC (DecontX score + neuronal marker):
obj <- flag_contamination(obj)                 # adds con_level / qc_flag
obj <- filter_contaminated(obj)                # then re-run NMF on the clean object
```

### Interpreting the modules

```{r nmf-interp, eval = FALSE}
tops <- nmf_top_pathways(nm, n = 30)                 # top pathways per factor
fbs  <- nmf_factor_by_state(obj, "state_anno2")      # factor x state mean loading
```

## Part 2 — The reduction engine (runnable)

Here we switch to simulated gene sets so the rest of the vignette executes with
only the CRAN dependencies. We build three redundancy "families" (A/B/C) that
each share a core gene set, plus one distinct set.

```{r simulate}
set.seed(1)
pool   <- paste0("g", 1:60)
core_A <- sample(pool, 8)
core_B <- sample(pool, 8)
core_C <- sample(pool, 8)
mk <- function(core, extra) unique(c(core, sample(pool, extra)))

genesets <- list(
  A1 = mk(core_A, 2), A2 = mk(core_A, 3), A3 = mk(core_A, 2),
  B1 = mk(core_B, 2), B2 = mk(core_B, 3), B3 = mk(core_B, 4),
  C1 = mk(core_C, 2), C2 = mk(core_C, 3),
  D1 = sample(pool, 10)
)
lengths(genesets)
```

### Collapse redundancy

`reduce_pathway_terms()` runs the whole standardize → similarity → group →
represent pipeline. With the overlap coefficient and a 0.5 threshold, the A/B/C
families should each collapse to a single representative while D1 stands alone.

```{r reduce}
res <- reduce_pathway_terms(
  genesets,
  similarity_method    = "overlap",
  reduction_method     = "hclust",
  similarity_threshold = 0.5,
  representative_rule  = "largest_geneset"
)

res$selected_terms
```

```{r reduce-table}
res$representative_table[, c("representative_term", "group_id",
                             "n_terms_in_group", "geneset_size")]
```

The `member_terms` column records which original pathways fell into each group:

```{r members}
setNames(res$representative_table$member_terms,
         res$representative_table$representative_term)
```

### Choosing representatives by an external score

This is exactly what the bridge does per state: instead of picking the largest
gene set, it supplies the NMF loading as a priority so the pathway with the
highest loading in that state wins its group. Here we mimic it with an arbitrary
priority vector via `representative_rule = "user_priority"`.

```{r priority}
# Pretend these are NMF loadings; A2 and B1 are favoured within their families.
priority <- c(A1 = 0.2, A2 = 0.9, A3 = 0.3,
              B1 = 0.8, B2 = 0.4, B3 = 0.5,
              C1 = 0.6, C2 = 0.7, D1 = 0.5)

res_prio <- reduce_pathway_terms(
  genesets,
  similarity_method    = "overlap",
  similarity_threshold = 0.5,
  representative_rule  = "user_priority",
  term_priority        = priority
)
res_prio$selected_terms   # A2 and B1 now represent their families
```

## Part 3 — Diagnostics and plots

### Threshold sensitivity

How many representatives survive as the threshold varies:

```{r sensitivity, fig.alt = "Number of retained representative terms across similarity thresholds"}
plot_reduction_sensitivity(
  genesets,
  thresholds        = seq(0.3, 0.9, by = 0.1),
  similarity_method = "overlap",
  reduction_method  = "hclust"
)
```

### Group-size distribution

```{r group-size, fig.alt = "Histogram of redundancy group sizes"}
plot_group_size_distribution(res)
```

### Similarity heatmap

Uses ComplexHeatmap or pheatmap when installed, otherwise a base heatmap.

```{r heatmap, fig.alt = "Pairwise gene-set similarity heatmap"}
plot_similarity_heatmap(res, max_terms = 50)
```

### Dendrogram with the cut height

Only meaningful for `reduction_method = "hclust"`; shown when `ggdendro` is
available.

```{r dendro, eval = requireNamespace("ggdendro", quietly = TRUE), fig.alt = "Redundancy dendrogram with threshold cut line"}
plot_dendrogram_cutoff(res)
```

### Graph mode

The graph reduction connects terms whose similarity exceeds the threshold and
takes connected components (or Louvain communities).

```{r graph, fig.alt = "Redundancy graph coloured by group"}
res_g <- reduce_pathway_terms(
  genesets,
  similarity_method    = "overlap",
  reduction_method     = "graph",
  community_detection  = "components",
  similarity_threshold = 0.5,
  representative_rule  = "largest_geneset"
)
plot_redundancy_graph(res_g)
```

## Part 4 — The bridge (on your Seurat object)

Once clusters are annotated into states (e.g. `obj$state_anno2`), the bridge
assigns NMF factors to states, gathers each state's top-loading pathways, and
reduces their gene-overlap redundancy — with the NMF loading picking each group's
representative.

```{r bridge, eval = FALSE}
bridge <- reduce_pathways_for_states(
  obj, genesets = gs, nmf_model = nm,
  group.by = "state_anno2",
  n_top    = 30,
  assign   = "argmax",          # each factor -> its most-active state
  similarity_method    = "overlap",
  similarity_threshold = 0.7
)

bridge$representative_table                        # per-state representatives + loading
plot_similarity_heatmap(bridge$results[["Antiviral"]])
```

Use `assign = "zscore"` instead if you want shared modules to contribute to
several states (a factor whose z-score exceeds `zscore_thresh` in more than one
state participates in each).

## One-call orchestration

`run_statecommute()` runs gene sets → scoring → NMF → clustering. Per-state
reduction is only attempted when you pass a `state_col`, since annotation is a
manual step in between.

```{r pipeline, eval = FALSE}
res_all <- run_statecommute(obj, k = 20, resolution = 0.3)
# ... annotate res_all$object$state_anno2 ...
bridge  <- reduce_pathways_for_states(
  res_all$object, genesets = res_all$genesets,
  nmf_model = res_all$nmf_model, group.by = "state_anno2"
)
```

## Session info

```{r session}
sessionInfo()
```
